\documentclass[12pt]{report}

% Packages.
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage[backend=biber]{biblatex}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{hyperref}
\usepackage[parfill]{parskip}
\usepackage{subfiles}
\usepackage{xcolor}

% Configuration.
\graphicspath{{images/}{../images/}}

\lstset{
  backgroundcolor=\color{white!99!black},
  numbers=left, numberstyle=\tiny, stepnumber=1, numbersep=5pt,
}

\lstdefinestyle{Caustic}{
  basicstyle=\small\ttfamily\color{black},
  breaklines=true,
  moredelim=[s][\color{green!50!black}\ttfamily]{'}{'},
  moredelim=*[s][\color{black}\ttfamily]{options}{\}},
  commentstyle={\color{gray}\itshape},
  morecomment=[l]{//},
  emph={def, del, elif, else, extends, if, import, module, record, return, rollback, service, while, var},
  emphstyle={\color{blue}\ttfamily},
  alsoletter={:,|,;},
  morekeywords={:,|,;},
  keywordstyle={\color{black}},
}

\hypersetup{
  colorlinks=true,
  linkcolor=black,
  citecolor=black,
  filecolor=blue,
  urlcolor=blue
}

\newtheorem{theorem}{Theorem}

% Bibliography.
\bibliography{caustic/caustic}
\bibliography{beaker/beaker}
\nocite{*}
\bibliographystyle{plain}

\begin{document}

% Title.
\title{On Transactional Programming}
\author{Ashwin Madavan}
\maketitle

% Abstract.
\begin{abstract}
  Concurrency is \emph{hard}. Concurrency refers to situations in which multiple programs
  simultaneously modify shared data. Concurrent programs may be run across threads, processes,
  and, in the case of distributed systems, networks. Concurrency is challenging because it
  introduces ambiguity in execution order, and it is precisely this ambiguity that cause a class of
  failures known as race conditions. Race conditions manifest themselves in subtle ways in
  concurrent systems, and they can often be difficult to detect and challenging to remove.

  Many programming languages provide fundamental abstractions such as locks, semaphores, and
  monitors to explicitly deal with race conditions. Some, like Rust \cite{rust}, go a step further
  and are able to statically detect race conditions between concurrent threads. But none, however,
  are able guarantee that race conditions will be completely eliminated from a distributed system.
  Distributed systems form the computing backbone of nearly every major technology from social
  networks to video streaming, but their intricate complexity coupled with the inability to detect
  race conditions makes designing them extremely error-prone.

  In this thesis, we will explore transactional programming and its application to building
  race-free distributed systems. We will first discuss Caustic, a transactional programming language
  that allows programs to be safely distributed without any explicit synchronization or
  coordination. We will show that Caustic requires the underlying storage system to satisfy a
  minimal contract. We will then discuss Beaker, a distributed, fault-tolerant database that
  efficiently implements this contract.
\end{abstract}

% Table of Contents.
\tableofcontents

% Caustic.
\chapter{Caustic: A Transactional Programming Language}
\subfile{caustic/caustic}

% Beaker.
\chapter{Beaker: A Distributed, Transactional Database}
\subfile{beaker/beaker}

% Appendix.
\appendix
\chapter{Caustic}
\subfile{caustic/appendix}

\chapter{Beaker}
\subfile{beaker/appendix}

% References.
\printbibliography

\end{document}